Eye in a Disk: eyeIntegration human eye and body transcriptome database version 1.00

### 2019-01-20 manuscript.Rmd
captioner captions

---
title: 'Eye in a Disk: eyeIntegration human eye and body transcriptome database version 1.00'
author:
  - Vinay Swamy:
      institute: ogvfb
  - David McGaughey:
      institute:
        - ogvfb
      correspondence: "yes"
      email: mcgaugheyd@mail.nih.gov
institute:
  - ogvfb: Ophthalmic Genetics and Visual Function Branch, National Eye Institute, National Institutes of Health
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  bookdown::word_document2:
    reference_docx: word-styles-reference-01.docx
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
bibliography: eyeIntegration_v1_app_manuscript.bib
csl: investigative-ophthalmology-and-visual-science.csl
---

@@

# I usually load my libraries up front to keep things organized
#library(bookdown)
#library(knitr)
#library(kableExtra)
library(pool)
library(RSQLite)
library(tidyverse)
library(citr)
library(cowplot)
library(ggrepel)
library(colorspace)
library(png)

gene_pool_2019 <- dbPool(drv = SQLite(), dbname = '/Volumes/data/eyeIntegration_EiaD/eyeIntegration_human_expression_2019_v100.sqlite')
gene_pool_2017 <- dbPool(drv = SQLite(), dbname = '/Volumes/data/eyeIntegration_EiaD/eyeIntegration_human_2017_01.sqlite')
core_tight_2017 <- gene_pool_2017 %>% tbl('metadata') %>% as_tibble()
core_tight_2019 <- gene_pool_2019 %>% tbl('metadata') %>% as_tibble()
load('data/human_tx_studies.Rdata')
tsne_50 <- gene_pool_2019 %>% tbl('tSNE_bulk_RNA') %>% filter(Perplexity == 45) %>% as_tibble()
anno <- gene_pool_2019 %>% tbl('gene_IDs')

@@

From anterior to posterior, the human eye is composed of the cornea, lens, retina, and RPE (choroid). The development and function of these tissues is mediated through tissue and temporal specific transcript and gene expression patterns, or transcriptome. Today, RNA-sequencing (RNA-seq) is the predominant technology in quantification of the transcriptome. Analysis of gene patterns across tissue, time, and perturbation allows researchers to decipher the genetic controls of eye development and function. To this end researchers have used a wide variety of tissue types to assess gene function, including primary tissue (fetal and post-mortem), induced pluripotent stem cell (iPSC) derived cells, immortalized cell lines, and most recently, organoids. These tissue types have been deeply sequenced across the cornea [@chenIdentificationNovelMolecular2013a; @chngHighThroughputGene2013a; @chungTranscriptomicProfilingPosterior2017; @fraustoTranscriptomicAnalysisCultured2016a; @kabzaCollagenSynthesisDisruption2017; @ouyangWNT7APAX6Define2014a; @songDirectedDifferentiationHuman2016a], lens [@hanWnt5aContributesDifferentiation2018], retina [@aldiriDynamicEpigeneticLandscape2017a; @farkasTranscriptomeAnalysesHuman2013b; @hoshinoMolecularAnatomyDeveloping2017; @kaewkhawTranscriptomeDynamicsDeveloping2015; @kaewkhawTreatmentParadigmsRetinal2016; @liComprehensiveAnalysisGene2014a; @mustafiTranscriptomeAnalysisReveals2016; @pinelliAtlasGeneExpression2016; @whitmoreTranscriptomicAnalysisNasal2014b], and RPE (choroid) [@darrowDeletionDXZ4Human2016a; @harenzaTranscriptomicProfiling392017; @huIdentificationMiRNASignatures2012b; @liComprehensiveAnalysisGene2014a; @nozawaSAFARegulatesInterphase2017; @obersteinCellularResponsesHuman2017; @pengEngineeringBloodRetinalBarrier2013; @radekeRestorationMesenchymalRetinal2015a; @sainiNicotinamideAmelioratesDisease2017; @samuelAppropriatelyDifferentiatedARPE192017; @santaguidaAneuploidyinducedCellularStresses2015b; @shaoSpontaneousGenerationNovel2017; @shihRestorationMesenchymalRPE2017; @smithRetinalPigmentEpithelial2017; @stevensonGiantinknockoutModelsReveal2017; @tresiniCoreSpliceosomeTarget2015; @whewaySiRNAbasedFunctionalGenomics2015; @whitmoreTranscriptomicAnalysisNasal2014b; @auCharacterizationLincRNAExpression2017; @huIdentificationMiRNASignatures2012b].

@@

## eyeIntegration web app and R package
The data generated in the above steps is consolidated into a SQLite database, with the original EiaD data from 2017 and the new 2019 EiaD dataset each getting a separated database file. The code that creates the eyeIntegration web app is written in Shiny and R and has been wrapped into an R package (https://github.com/davemcg/eyeIntegration_app/) that can be deployed on a local computer or a web server (https://eyeIntegration.nei.nih.gov). The app can be deployed on a local computer with 40GB of free disk space by running three commands in R: "devtools::install_github('davemcg/eyeIntegration_app'), eyeIntegrationApp::get_eyeIntegration_datasets(), eyeIntegrationApp::run_eyeIntegration()."

@@

## scRNA-seq processing
The eyeIntegration site currently hosts two large scRNA-seq datasets from Macoscko et al. and a unpublished set kindly made available to the public from Clark et al [@clarkComprehensiveAnalysisRetinal2018; @macoskoHighlyParallelGenomewide2015a]. We use the processed gene count data from each, as well as the cluster assignments that specify what cell type each individual cell is. The count data is mean averaged to the cell type, age, and gene level. We also display t-SNE and UMAP-based two-dimensional visualizations of the Macosko and Clark data, respectively, in the web app. For more details see Methods of [@Bryan_2018].

@@

## Rigorous quality control and reproducible workflow system ensures high quality transcriptomes

We built an automated pipeline for processing and analyzing all data for the web app using the program Snakemake, a python-based workflow management system that allows for efficient parallel execution of the analysis, facilitates reproduction by others, and simplifies long term maintenance of the EiaD data [@kosterSnakemakeScalableBioinformatics2012]. In order to create a high quality final dataset, we developed a rigorous quality control procedure as part of our analysis, considering a sample’s read mapping rate and median count level as well as behavior relative to samples of the same sub-tissue type (see Methods).  To identify possible tube switches or sample extraction issues, we used t-SNE to group samples by sub-tissue type to identify variability within samples of the same sub-tissue and ensure overall consistency in data processing (Figure \@ref(fig:tSNE)).

@@

```{r tSNE, fig.height = 15, fig.width = 20, echo=F, message = F, fig.cap="Tissue - tissue consistent two-dimensional sample - sample transcriptome profiles demonstrate effective quality control and sample processing"}
dbscan_cluster <- dbscan::dbscan(tsne_50 %>% select(X1, X2),  minPts = 3, eps = 2.5)
tsne_50$Cluster <- dbscan_cluster$cluster
col1 = 'X1'
col2 = 'X2'
# create label points for each cluster
cluster_centers <- tsne_50 %>%
  left_join(.,core_tight_2019) %>% group_by(Cluster) %>%
  summarise(C1=mean(!!sym(col1)),C2=mean(!!sym(col2)),Tissue=paste(unique(Tissue),collapse=','))

@@

## `r ((core_tight_2019$study_accession %>% unique()) %in% (core_tight_2017$study_accession %>% unique())) %>% table() %>% as_tibble() %>% slice(1) %>% pull(n)` new studies and `r ((core_tight_2019 %>% filter(Kept=='Kept', study_accession != 'SRP012682') %>% pull(sample_accession)) %>% length()) - core_tight_2017 %>% filter(study_accession != 'SRP012682') %>% nrow()` new samples, and four new sub-tissue types added to 2019 EiaD dataset
Our initial query to the SRA found `r human_tx_studies$study_accession %>% unique() %>% length()` potentially relevant studies deposited since our last search on January 19th, 2017. We removed non-pertinent studies and selected healthy or unmodified tissue from each relevant study for a total, including  of `r core_tight_2019$study_accession %>% unique() %>% length()` studies, `r ((core_tight_2019$study_accession %>% unique()) %in% (core_tight_2017$study_accession %>% unique())) %>% table() %>% as_tibble() %>% slice(1) %>% pull(n)` of which are new to the 2019 EiaD dataset. The 2019 EiaD dataset contains `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession != 'SRP012682') %>% nrow()` human eye tissues and also includes `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession == 'SRP012682') %>% nrow()` GTEx samples across `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession == 'SRP012682') %>% pull(Sub_Tissue) %>% table() %>% length()` tissues for easy comparison. The 2019 EiaD contains `r core_tight_2019 %>% filter(Kept == 'Kept', Tissue == 'Cornea') %>% nrow()` cornea, `r core_tight_2019 %>% filter(Kept == 'Kept', Tissue == 'Lens') %>% nrow()` lens, `r core_tight_2019 %>% filter(Kept == 'Kept', Tissue == 'Retina') %>% nrow()` retina, and `r core_tight_2019 %>% filter(Kept == 'Kept', Tissue == 'RPE') %>% nrow()` RPE (choroid) samples; in total we have added `r ((core_tight_2019 %>% filter(Kept=='Kept', study_accession != 'SRP012682') %>% pull(sample_accession)) %>% length()) - core_tight_2017 %>% filter(study_accession != 'SRP012682') %>% nrow()` new samples to the 2019 EiaD (Figure \@ref(fig:tissueCountsBarPlot)). Stem cell derived cornea, stem cell derived lens, fetal retina, and organoid retina are four new types of sub-tissues that are now available in eyeIntegration.
Another substantial addition to the 2019 EiaD are non-protein coding genes. We now quantify expression across `r gene_pool_2019 %>% tbl('gene_IDs') %>% count(gene_type) %>% as_tibble() %>% nrow()` gene types and  `r gene_pool_2019 %>% tbl('tx_IDs') %>% count(transcript_type) %>% as_tibble() %>% nrow()` transcript types, including protein coding, retained intron, lincRNA, antisense, and pseuodgenes.

```{r tissueCountsBarPlot, fig.height = 4, fig.cap="Comparison of counts of eye tissue samples from 2017 to 2019 EiaD", echo=F}
core_both <- bind_rows(core_tight_2019 %>% mutate(Version = '2019 EiaD') %>% select(-run_accession) %>% filter(Kept == 'Kept'), core_tight_2017 %>% mutate(Version = '2017') %>% mutate(Sub_Tissue = gsub('_',' - ', Sub_Tissue))) %>% unique()

@@

## eyeIntegration web app provides interactive visual portal to all data 
All data generated by our analyses is presented through the eyeIntegration web app (https://eyeIntegration.nei.nih.gov). The web app was designed to provide a simple interface that have the same general concept – select specific genes and tissue and view relevant information. The web-app is divided into four general categories: expression, two-dimension sample (or single cell) relationships, gene networks, and data tables.

@@

## Gene expression in individual retina cells types with scRNA-seq 
We provide scRNA-seq data from murine retina across two studies [@clarkComprehensiveAnalysisRetinal2018; @macoskoHighlyParallelGenomewide2015a], allowing for examination of gene expression in individual cell types in the retina. The count data is taken directly from the publishing authors (see Methods). Single cell gene expression data is visualized through a heatmap showing the expression of a single gene across multiple retinal cell types and different developmental time points (E11-P14), and a table of expression values is generated containing the expression data used to draw the heatmap (Figure \@ref(fig:appScreenshots)C). We also provide t-SNE/UMAP based clustering using cell type specific labeling created by the publishing authors (Figure \@ref(fig:appScreenshots)D). The plots show all cell types present a given developmental stage, and highlights cells expressing a gene at a user determined level.

>>>>

---
title: 'Eye in a Disk: eyeIntegration human eye and body transcriptome database version 1.00'
author:
  - Vinay Swamy:
      institute: ogvfb
  - David McGaughey:
      institute:
        - ogvfb
      correspondence: "yes"
      email: mcgaugheyd@mail.nih.gov
institute:
  - ogvfb: Ophthalmic Genetics and Visual Function Branch, National Eye Institute, National Institutes of Health
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  word_document:
    reference_docx: word-styles-reference-01.docx
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
bibliography: eyeIntegration_v1_app_manuscript.bib
csl: investigative-ophthalmology-and-visual-science.csl
---

@@

library(pool)
library(RSQLite)
library(tidyverse)
library(citr)
library(cowplot)
library(ggrepel)
library(colorspace)
library(png)
library(flextable)
library(captioner)

# setup caption-ing
fig_cap <- captioner("Figure")
supFig_cap <- captioner("Supplemental Figure")
tab_cap <- captioner("Table")
supTab_cap <- captioner("Supplemental Table")

# attach data pools
gene_pool_2019 <- dbPool(drv = SQLite(), dbname = '/Volumes/data/eyeIntegration_EiaD/eyeIntegration_human_expression_2019_v100.sqlite')
gene_pool_2017 <- dbPool(drv = SQLite(), dbname = '/Volumes/data/eyeIntegration_EiaD/eyeIntegration_human_2017_01.sqlite')
core_tight_2017 <- gene_pool_2017 %>% tbl('metadata') %>% as_tibble()
core_tight_2019 <- gene_pool_2019 %>% tbl('metadata') %>% as_tibble()
load('data/human_tx_studies.Rdata')
tsne_50 <- gene_pool_2019 %>% tbl('tSNE_bulk_RNA') %>% filter(Perplexity == 45) %>% as_tibble()
anno <- gene_pool_2019 %>% tbl('gene_IDs')

@@

From anterior to posterior, the human eye is composed of the cornea, lens, retina, and RPE (choroid). The development and function of these tissues is mediated through tissue and temporal specific transcript and gene expression patterns, or transcriptome. Today, RNA-sequencing (RNA-seq) is the predominant technology in quantification of the transcriptome. Analysis of gene patterns across tissue, time, and perturbation allows researchers to decipher the genetic controls of eye development and function. To this end researchers have used a wide variety of tissue types to assess gene function, including primary tissue (fetal and post-mortem), induced pluripotent stem cell (iPSC) derived cells, immortalized cell lines, and most recently, organoids. These tissue types have been deeply sequenced across the cornea [@chenIdentificationNovelMolecular2013a; @chngHighThroughputGene2013a; @chungTranscriptomicProfilingPosterior2017; @fraustoTranscriptomicAnalysisCultured2016a; @kabzaCollagenSynthesisDisruption2017; @ouyangWNT7APAX6Define2014a; @songDirectedDifferentiationHuman2016a], lens [@hanWnt5aContributesDifferentiation2018], retina [@aldiriDynamicEpigeneticLandscape2017a; @farkasTranscriptomeAnalysesHuman2013b; @hoshinoMolecularAnatomyDeveloping2017; @kaewkhawTranscriptomeDynamicsDeveloping2015; @kaewkhawTreatmentParadigmsRetinal2016; @liComprehensiveAnalysisGene2014a; @mustafiTranscriptomeAnalysisReveals2016; @pinelliAtlasGeneExpression2016; @whitmoreTranscriptomicAnalysisNasal2014b], and RPE (choroid) [@darrowDeletionDXZ4Human2016a; @harenzaTranscriptomicProfiling392017; @huIdentificationMiRNASignatures2012b; @liComprehensiveAnalysisGene2014a; @nozawaSAFARegulatesInterphase2017; @obersteinCellularResponsesHuman2017; @pengEngineeringBloodRetinalBarrier2013; @radekeRestorationMesenchymalRetinal2015a; @sainiNicotinamideAmelioratesDisease2017; @samuelAppropriatelyDifferentiatedARPE192017; @santaguidaAneuploidyinducedCellularStresses2015b; @shaoSpontaneousGenerationNovel2017; @shihRestorationMesenchymalRPE2017; @smithRetinalPigmentEpithelial2017; @stevensonGiantinknockoutModelsReveal2017; @tresiniCoreSpliceosomeTarget2015; @whewaySiRNAbasedFunctionalGenomics2015; @whitmoreTranscriptomicAnalysisNasal2014b; @auCharacterizationLincRNAExpression2017].

@@

## eyeIntegration web app and R package
The data generated in the above steps is consolidated into a SQLite database, with the original EiaD data from 2017 and the new 2019 EiaD dataset each getting a separated database file. The code that creates the eyeIntegration web app is written in Shiny and R and has been wrapped into an R package (https://github.com/davemcg/eyeIntegration_app/) that can be deployed on a local computer or a web server (https://eyeIntegration.nei.nih.gov). The app can be deployed on a local computer with 40GB of free disk space by running three commands in R: "devtools::install_github('davemcg/eyeIntegration_app')", "eyeIntegrationApp::get_eyeIntegration_datasets()", and  "eyeIntegrationApp::run_eyeIntegration()".

@@

## scRNA-seq processing
The eyeIntegration site currently hosts two large scRNA-seq datasets from Macoscko et al. and a unpublished set kindly made available to the public from Clark et al [@clarkComprehensiveAnalysisRetinal2018; @macoskoHighlyParallelGenomewide2015a]. We use the processed gene count data from each group, as well as the cluster assignments that specify what cell type each individual cell is. The count data is mean averaged to the cell type, age, and gene level for the single cell expression section of eyeIntegration. We also display t-SNE and UMAP-based two-dimensional visualizations of the Macosko and Clark data, respectively, in the web app. For detail so the t-SNE processing we did on the Macosko dataset, see the methods of [@Bryan_2018].

@@

## Rigorous quality control and reproducible workflow system ensures high quality transcriptomes
```{bash, echo = F, eval = F}
# ran in biowulf2.nih.gov:/data/mcgaugheyd/projects/nei/mcgaughey/auto_eyeIntegration
# 1314 from running: for i in RE_quant_files/*/lib_format_counts.json; do echo $i; done | wc -l 
# 46030412878 from running: for i in RE_quant_files/*/lib_format_counts.json; do grep num_compatible_fragments $i; done |  cut -f2 -d":" | cut -f1 -d"," | awk '{sum += $1} END {print sum}'
```
We built an automated pipeline for processing and analyzing all data for the web app using the program Snakemake, a python-based workflow management system that allows for efficient parallel execution of the analysis, facilitates reproduction by others, and simplifies long term maintenance of the EiaD data [@kosterSnakemakeScalableBioinformatics2012]. In order to create a high quality final dataset across the 1,314 initial samples and 46,030,412,878 reads we developed a rigorous quality control procedure as part of our analysis, considering a sample’s read mapping rate and median count level as well as behavior relative to samples of the same sub-tissue type (see Methods).  To identify tube switches or sample extraction issues, we used t-SNE to group samples by sub-tissue type to identify variability within samples of the same sub-tissue and ensure overall consistency in data processing (`r fig_cap(name = 'tSNE', display = 'cite')`).

@@

```{r tsne_fig, fig.height = 15, fig.width = 20, echo = F, message = F, fig.cap=tsne_cap}
tsne_cap <- fig_cap(name='tSNE', caption = 'Tissue - tissue consistent two-dimensional sample - sample transcriptome profiles demonstrate effective quality control and sample processing')

dbscan_cluster <- dbscan::dbscan(tsne_50 %>% select(X1, X2),  minPts = 3, eps = 2.5)
tsne_50$Cluster <- dbscan_cluster$cluster
col1 = 'X1'
col2 = 'X2'
# create label points for each cluster
cluster_centers <- tsne_50 %>%
  left_join(.,core_tight_2019) %>% group_by(Cluster) %>%
  summarise(C1=mean(!!sym(col1)),C2=mean(!!sym(col2)),Tissue=paste(unique(Tissue),collapse=','))

@@

## EiaD 2019 contains `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession == 'SRP012682') %>% pull(Sub_Tissue) %>% table() %>% length()` GTEx body sample along with `r ((core_tight_2019$study_accession %>% unique()) %in% (core_tight_2017$study_accession %>% unique())) %>% table() %>% as_tibble() %>% slice(1) %>% pull(n)` new studies and `r ((core_tight_2019 %>% filter(Kept=='Kept', study_accession != 'SRP012682') %>% pull(sample_accession)) %>% length()) - core_tight_2017 %>% filter(study_accession != 'SRP012682') %>% nrow()` new samples, and four new sub-tissue types added to 2019 EiaD dataset
Our initial query to the SRA found `r human_tx_studies$study_accession %>% unique() %>% length()` potentially relevant studies deposited since our last search on January 19th, 2017. We removed non-pertinent studies and selected healthy or unmodified tissue from each relevant study for a total, including  of `r core_tight_2019$study_accession %>% unique() %>% length()` studies, `r ((core_tight_2019$study_accession %>% unique()) %in% (core_tight_2017$study_accession %>% unique())) %>% table() %>% as_tibble() %>% slice(1) %>% pull(n)` of which are new to the 2019 EiaD dataset. The 2019 EiaD dataset contains `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession != 'SRP012682') %>% nrow()` human eye tissues and also includes `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession == 'SRP012682') %>% nrow()` GTEx samples across `r core_tight_2019 %>% filter(Kept == 'Kept') %>% filter(study_accession == 'SRP012682') %>% pull(Sub_Tissue) %>% table() %>% length()` tissues for easy comparison `r paste0('Tables ', tab_cap(name = 'tableSampleCounts', display = 'num'), ', ', tab_cap(name = 'tableSampleCountsGTEx', display = 'num'))`. The 2019 EiaD contains `r core_tight_2019 %>% filter(Kept == 'Kept', Tissue == 'Cornea') %>% nrow()` cornea, `r core_tight_2019 %>% filter(Kept == 'Kept', Tissue == 'Lens') %>% nrow()` lens, `r core_tight_2019 %>% filter(Kept == 'Kept', Tissue == 'Retina') %>% nrow()` retina, and `r core_tight_2019 %>% filter(Kept == 'Kept', Tissue == 'RPE') %>% nrow()` RPE (choroid) samples; in total we have added `r ((core_tight_2019 %>% filter(Kept=='Kept', study_accession != 'SRP012682') %>% pull(sample_accession)) %>% length()) - core_tight_2017 %>% filter(study_accession != 'SRP012682') %>% nrow()` new samples to the 2019 EiaD  (`r fig_cap(name = 'tissueCountsBarPlot', display = 'cite')`). Stem cell derived cornea, stem cell derived lens, fetal retina, and organoid retina are four new types of sub-tissues that are now available in eyeIntegration.
Another substantial addition to the 2019 EiaD are non-protein coding genes. We now quantify expression across `r gene_pool_2019 %>% tbl('gene_IDs') %>% count(gene_type) %>% as_tibble() %>% nrow()` gene types and  `r gene_pool_2019 %>% tbl('tx_IDs') %>% count(transcript_type) %>% as_tibble() %>% nrow()` transcript types, including protein coding, retained intron, lincRNA, antisense, and pseuodgenes.

```{r tissueCountsBarPlot, fig.height = 4, fig.cap=tissueCountsBarPlot_cap, echo=F}
tissueCountsBarPlot_cap <- fig_cap(name='tissueCountsBarPlot', caption = 'Substantial increase in eye tissue count and type  from 2017 to 2019 EiaD')

core_both <- bind_rows(core_tight_2019 %>% mutate(Version = '2019 EiaD') %>% select(-run_accession) %>% filter(Kept == 'Kept'), core_tight_2017 %>% mutate(Version = '2017') %>% mutate(Sub_Tissue = gsub('_',' - ', Sub_Tissue))) %>% unique()

@@

```{r tableSampleCounts, echo = F, results='asis', warning = F, fig.cap = tableSampleCounts_cap}
tableSampleCounts_cap <- tab_cap(name='tableSampleCounts', caption = 'EiaD and eyeIntegration contain a huge set of eye tissues')
table <- left_join(core_tight_2019  %>% 
                     select(-run_accession) %>% 
                     unique() %>% 
                     filter(study_accession != 'SRP012682',Tissue %in% c('RPE','Lens','Cornea', 'Retina')) %>% 
                     group_by(Tissue) %>% 
                     summarise(`Pre QC Count` = n()),

                   core_tight_2019  %>% 
                     select(-run_accession) %>% 
                     unique() %>% 
                     filter(study_accession != 'SRP012682', Tissue %in% c('RPE','Lens','Cornea', 'Retina'), Kept == 'Kept') %>% 
                     rowwise() %>% 
                     mutate(Sub = str_split(Sub_Tissue, ' - ')[[1]][2]) %>% 
                     mutate(Sub = case_when(is.na(Sub) ~ Sub_Tissue, TRUE ~ Sub)) %>% 
                     group_by(Tissue, Sub) %>% summarise(Count = n()) %>% 
                     mutate(Sub = paste0(Sub, ' (', Count, ')')) %>% 
                     ungroup() %>% 
                     group_by(Tissue) %>% 
                     summarise(`Count` = sum(Count), `Sub-Tissue Types (Count)` = paste(sort(unique(trimws(Sub))), collapse=', ')), by = 'Tissue') %>% 
  as_tibble()

gtex_tissues <- core_tight_2019 %>% 
  select(-run_accession) %>% 
  unique() %>%
  filter(study_accession == 'SRP012682', Kept == 'Kept') %>%
  group_by(Tissue, study_accession) %>% 
  summarise(Count = n()) %>% 
  ungroup() %>% 
  mutate(Tissue = paste0(Tissue, ' (', Count, ')')) %>% 
  pull(Tissue)

gtex <- cbind('GTEx', core_tight_2019 %>% 
                select(-run_accession) %>% 
                unique() %>%
                filter(study_accession == 'SRP012682') %>% nrow(),
              core_tight_2019 %>% 
                select(-run_accession) %>% 
                unique() %>%
                filter(study_accession == 'SRP012682', Kept == 'Kept') %>% nrow(),
              paste(gtex_tissues, collapse = ', ')) %>% as_tibble()

colnames(gtex) <- c('Group','Pre QC Count', 'Count', 'Tissue Types (Count)')
cat("<div id=\"tbl:tableSampleCounts\">")
regulartable(table) %>% width(width = c(0.5,1,0.5, 2)) %>% add_footer(., tableSampleCounts_cap) %>% flextable::align(align='center')
#Output Caption
cat(paste0("\n",tableSampleCounts_cap,"\n"))
# Output </div>
cat("</div>")
```
```{r tableSampleCountsGTEx, echo = F, results='asis', warning = F, fig.cap = tableSampleCountsGTEx_cap}
tableSampleCountsGTEx_cap <- tab_cap(name='tableSampleCountsGTEx', caption = 'EiaD holds hundreds of GTEx tissues to provide a broad comparison set')
cat("<div id=\"tbl:tableSampleCountsGTEx\">")
regulartable(gtex) %>% width(width = c(0.5,1,0.5,5)) %>% flextable::align(align='center')
#Output Caption
cat(paste0("\n",tableSampleCountsGTEx_cap,"\n"))
# Output </div>
cat("</div>")
```

## eyeIntegration web app provides interactive visual portal to all data 
The EiaD 2019 dataset is presented through the eyeIntegration web app (https://eyeIntegration.nei.nih.gov). The web app was designed to provide a simple interface that have the same general concept – select specific genes and tissue and view relevant information. The web-app is divided into four general categories: expression, two-dimension sample (or single cell) relationships, gene networks, and data tables.

@@

## Gene expression in individual retina cells types with scRNA-seq 
We provide scRNA-seq data from murine retina across two studies [@clarkComprehensiveAnalysisRetinal2018; @macoskoHighlyParallelGenomewide2015a]. This allows researchers to quickly examine gene expression across individual cell types in the retina. Single cell gene expression data is visualized through a heatmap showing the expression of a gene across multiple retinal cell types and different developmental time points (E11-P14, if available), and a table of expression values is generated containing the expression data used to draw the heatmap (Figure \@ref(fig:appScreenshots)C). We also provide t-SNE/UMAP based clustering using cell type specific labeling created by the publishing authors (Figure \@ref(fig:appScreenshots)D, see Methods). The plots show all cell types present a given developmental stage, and highlights cells expressing a gene above a user given level.

## Retina Organoids, Fetal Tissue, and iPSC-derived retina are closely related

There are, currently, three approaches to collecting human fetal retina: 1. post-mortem tissue, 2. organoids, and 3. iPSC-derived cells. As an example of how the EiaD 2019 dataset can be used to answer pan-study questions, we looked at whether at how well these three approaches to generating fetal retina compare at a transcriptomic level. Should organoid and iPSC-derived cell line retina compare closely to fetal tissue, this would further re-inforce the value of these approaches in studying human retina development.
